Binning Strategies and Related Loss for Large Data

### 2015-03-31 Paper.Rnw
Removing proportion of potential loss. Scaling by worst-case scenario was more of a headache than it was worth.  Comparisons of net spatial loss are sufficient.  Figure 8 much more interpretable now.

The net spatial loss is a euclidean distance, but requires a reference scale on which the magnitude may be interpretable. We  use the spatial loss from a ``worst-case'' binning scenario as a reference value. The largest displacement of points  occurs if we place all points into a single bin.  While the total spatial loss for single bin could be made arbitrarily large if the bin center is placed far away from the data values, we will consider a reasonable specification of a bin with boundaries that span only the range of the data in each dimension centered at  $( (x_{(1)}-x_{(n)})/2$ , $(y_{(1)}-y_{(n)})/2 )$.  The total spatial loss for this ``worst-case'' bin will be referred to as the \textit{potential spatial loss}, $L^S_p$, which is defined as 
\begin{eqnarray}\label{potspatloss}
L^S_p = \sum_{i=1}^{n} \sqrt{ \left(x_i-(x_{(1)}-x_{(n)})/2\right)^2 + \left(y_i-(y_{(1)}-y_{(n)})/2\right)^2 }
\end{eqnarray} 
%
We use net spatial loss as a proportion of potential spatial loss, $L^S_{net}/L^S_p$,  as the primary metric for assessing spatial loss incurred from binning.

@@

 Similar to spatial binning, \hh{we define} a worst-case scenario for frequency binning that provides context to the magnitude of the total frequency loss.  If we use a single frequency bin then all tiles of the binned scatterplot are rendered with the same shade; thus the shade  represents the average bin density \hh{XXX this is different from the spatial loss - visually the shade would be half way between lowest and highest frequency, wouldn't it?}, $\bar{c} =  (\sum_{\ell=1}^{L} c_\ell) /L$.  We define the total frequency loss in the single shade scenario as the \textit{potential frequency loss}, $L^F_p$,  where
%
\begin{eqnarray}\label{potfreqloss}
L^F_p = \sum_{\ell=1}^{L} (c_\ell-\bar{c})^2.
\end{eqnarray}  
%
Thus total frequency loss after binning can be interpreted through the proportion of potential frequency loss, $L^F / L^F_p$.  Similarly we define the \textit{potential log frequency loss}, $L^{\log F}_p$, as 
%
\begin{eqnarray}\label{potlogfreqloss}
L^{\log F}_p = \sum_{\ell=1}^{L} (\log(c_\ell+1)-\log(\bar{c}+1))^2,
\end{eqnarray}  
% 
which can then be used to find the proportion of potential log frequency loss, $L^{\log F} / L^{\log F}_p$, in cases where log frequency binning was conducted. \hh{XXX could you give a couple more details what you mean by cases where frequency binning was conducted?}

@@

These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places. Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles. To observe binning loss under more realistic conditions, we created three  data sets by rounding the values from the originally simulated data sets to the nearest even number. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.

@@

Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
Obviously, the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases. The proportion of potential spatial loss from random binning increases with bin width at a faster rate than from standard binning, as indicated by the widening gap between the lines and the steeper slopes.

@@

<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for proportion of potential spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>= 
spatlossdat <- read.csv("SpatialLossData.csv",header=TRUE)
# use all the square bins to plot bin width against proportion of max spataial loss
spatlossdatSquareBins <- spatlossdat[spatlossdat$widthx==spatlossdat$widthy,]
qplot(widthx,propSpatialLoss,geom="path",colour=bintype, linetype=bintype,
      data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
  theme_bw() + ylab("Prop. of Potential Spatial Loss") +
  xlab("Width/Height of Square Bins")
@

>>>>

% The net spatial loss is a euclidean distance, but requires a reference scale on which the magnitude may be interpretable. We  use the spatial loss from a ``worst-case'' binning scenario as a reference value. The largest displacement of points  occurs if we place all points into a single bin.  While the total spatial loss for single bin could be made arbitrarily large if the bin center is placed far away from the data values, we will consider a reasonable specification of a bin with boundaries that span only the range of the data in each dimension centered at  $( (x_{(1)}-x_{(n)})/2$ , $(y_{(1)}-y_{(n)})/2 )$.  The total spatial loss for this ``worst-case'' bin will be referred to as the \textit{potential spatial loss}, $L^S_p$, which is defined as 
% \begin{eqnarray}\label{potspatloss}
% L^S_p = \sum_{i=1}^{n} \sqrt{ \left(x_i-(x_{(1)}-x_{(n)})/2\right)^2 + \left(y_i-(y_{(1)}-y_{(n)})/2\right)^2 }
% \end{eqnarray} 
% %
% We use net spatial loss as a proportion of potential spatial loss, $L^S_{net}/L^S_p$,  as the primary metric for assessing spatial loss incurred from binning.

@@

%  Similar to spatial binning, \hh{we define} a worst-case scenario for frequency binning that provides context to the magnitude of the total frequency loss.  If we use a single frequency bin then all tiles of the binned scatterplot are rendered with the same shade; thus the shade  represents the average bin density \hh{XXX this is different from the spatial loss - visually the shade would be half way between lowest and highest frequency, wouldn't it?}, $\bar{c} =  (\sum_{\ell=1}^{L} c_\ell) /L$.  We define the total frequency loss in the single shade scenario as the \textit{potential frequency loss}, $L^F_p$,  where
% \begin{eqnarray}\label{potfreqloss}
% L^F_p = \sum_{\ell=1}^{L} (c_\ell-\bar{c})^2.
% Thus total frequency loss after binning can be interpreted through the proportion of potential frequency loss, $L^F / L^F_p$.  Similarly we define the \textit{potential log frequency loss}, $L^{\log F}_p$, as 
% \begin{eqnarray}\label{potlogfreqloss}
% L^{\log F}_p = \sum_{\ell=1}^{L} (\log(c_\ell+1)-\log(\bar{c}+1))^2,
% which can then be used to find the proportion of potential log frequency loss, $L^{\log F} / L^{\log F}_p$, in cases where log frequency binning was conducted. \hh{XXX could you give a couple more details what you mean by cases where frequency binning was conducted?}

@@

These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places. Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles. The data resolution can be described by increment between data values. To observe binning loss under more realistic conditions, we created three data sets by rounding the values from the originally simulated data sets to the nearest even number, thus a data resolution of 2 units. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.

@@

Figure~\ref{fig:SpatialLossCompare} displays the net spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$. It is easily apparent that  the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases. The net spatial loss from random binning increases with bin width at a faster rate than from standard binning, as indicated by the widening gap between the lines and the steeper slopes.

@@

<<SpatialLossCompare,echo=F,include=T,eval=T,fig.width=8, fig.height=3.5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Lineplots for net spatial loss over a range of bin sizes for standard and random binning of the fine version of the simulated data from each bivariate distribution.">>= 
spatlossdat <- read.csv("SpatialLossData.csv",header=TRUE)
# use all the square bins to plot bin width against proportion of max spataial loss
spatlossdatSquareBins <- spatlossdat[spatlossdat$widthx==spatlossdat$widthy,]
qplot(widthx,totalSpatialLoss,geom="path",colour=bintype, linetype=bintype,
      data=spatlossdatSquareBins,group=bintype, facets=.~distrib, size=I(1)) +
  theme_bw() + ylab("Net Spatial Loss") +
  xlab("Width/Height of Square Bins")
@
